Estimating the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations

### 2022-04-20 manuscript-Ghana/manuscript-Ghana.Rmd
Big updates

---
title: Understanding the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations
author:
  - name: Issoufou Liman Harou
    email: issoufoul@gmail.com
    affiliation: ICRAFGOV,KUENV
    correspondingauthor: true
  - name: Julliet Inyele 
    email: inyelejuli@gmail.com
    affiliation: KUGEO
  - name: Peter Minang  
    email: A.Minang@cgiar.org
    affiliation: ICRAFGOV
  - name: Lalisa Duguma
    email: l.duguma@cgiar.org
    affiliation: ICRAFGOV
address:
  - code: ICRAFGOV
    address: "Governance and Landscapes, World Agroforestry (ICRAF), UN Avenue, Gigiri, Nairobi, Kenya, P.O.Box 30677-00100"
  - code: KUENV
    address: "Department of Environmental Sciences, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O Box: 43844-00100"
  - code: KUGEO
    address: "Department of Geography, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O Box: 43844-00100"
abstract: |
  Remote sensing analysis and narratives reported evidence of mangrove degradation at various spatial scale across Ghana. Despite the importance of mangrove forests, the coverage of earth observation satellites, and the availability of cloud computing platforms for remote sensing data analysis, accurate country-wide information on the dynamics of these mangrove forests is lacking. Cloud cover is an important limitation for optical remote sensing in Southern Ghana. To understand the dynamics of mangrove forests, we identified 17 land cover classes, harmonized the Landsat scenes across sensors, used time series analysis to improve cloud detection and recover the relevant signal encoding the temporal signature of land cover. Based on the seasonal behavior of the land cover classes, the persistence of cloud cover limiting the number of usable images and the temporal resolution of Landsat satellite, we mosaiced the images within 30 days of their acquisition dates. Using the fitted image time series, we computed unsupervised classifications which we used along with ground control polygons to generate training and validation data for land cover classifications and change detection. The analysis of the root mean square error indicates that the accuracy of the fitted image time series increases with band wavelength and land cover seasonality. The accuracy assessment indicates an overall accuracy between 0.94 and 0.95 and a Kappa between 0.93 and 0.95 for the different land cover classifications. Mangrove forests have declined in some areas, whereas they have increased in others. Mangrove area, which has decreased between 2000 and 2010, has increased between 2010 and 2020. The gains experienced between 2010 and 2020, however, did not compensate for the lost the land cover experienced between 2000 and 2010. We identified four top most important land cover involved in mangroves conversion including forests with a high rate, croplands and natural vegetation mosaics and tree plantations with moderate rate and wetlands with a low rate. Better mangrove restoration and conservation are required to fully reverse the trend of mangrove decline.
keywords: 
  - Google Earth Engine
  - Land Cover Classification
  - Cloud Cover
  - Mangrove
  - Change Detection
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: bibliography.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/remote-sensing-of-environment
csl: remote-sensing-of-environment.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: natbib
link-citations: yes
linkcolor: blue
header-includes:
  - \input{includes.tex}
note: Replication files are available on the author's Github account (http://github.com/Issoufou-Liman).
---

@@

Despite the importance of mangroves, some studies reported that mangrove vegetation cover has been alarmingly declining in many areas of Ghana for the last few decades [@Asante-et-al-2017]. @Armah-et-al-2009 estimated this decline to range between 20 and 30% across the coast of Atlantic Ocean over the past 3 decades. @FAO-2007 reported a mangrove vegetation loss of 0.5 million ha (13.8 % of the total mangrove areas) in 25 years. According to @Rubin-et-al-1999, almost 2/3 of the Volta estuary mangroves have been lost since 1973, leaving only a few patches within a buffer of 15 km. @Darkwa-and-Smardon-2010 noticed an increased cutting of mangroves vegetation in Fosu Lagoon. @FAO-2007 estimated the decline of mangrove areas in Ghana at 24% between 1980 to 2005.

In other areas, however, other studies reported an appreciable increase in mangrove vegetation cover following restoration efforts by the locals. @Awo-et-al-2014 reported a noticeable increase in mangrove cover between 1986 and 2002 in Anyanui (Keta Lagoon). @Feka-2015 noted that despite the reduction in mangrove area in Ghana, an estimated area of 68 ha of mangroves have been restored from 1980 to 2006.

It becomes clear from the available information that a country-wide assessment of mangrove vegetation in Ghana is needed. A holistic assessment of all major Land Use Land Cover (LULC) would help to understand the dynamics of mangrove vegetation and guide mangrove conservation policy in the study area. This would help identify areas of increase and decrease in mangrove vegetation, understand the trend in LULC change, identify areas that require immediate action and those areas where policy interventions can achieve the greatest impact.

To our knowledge, country-wide assessments of mangrove vegetation that deliver this crucial information for mangrove conservation are lacking. The most recent and comprehensive ones (e.g., @CILSS-2016; @Spalding-et-al-1997; @Spalding-et-al-2010) have either coarse spatial resolution or have important information gap to be suitable for this purpose. This information gap is mainly due to the prevalence of cloud cover [@Ashiagbor-et-al-2021] that makes it virtually impossible to derive a complete land cover map from a single acquisition [@Rubin-et-al-1999].

Clouds deserved a standalone class in @CILSS-2016 because none of the best available image was cloud-free enough to allow a comprehensive post-classification interpolation. Based on our analysis of clouds identified by Fmask algorithm, at least half of the available Landsat images acquired between 2000 and 2020 presents missing observations over the coastal regions of Ghana. There is a need for approaches that could comprehensively account for these gaps when working with optical remote sensing data which are currently the best available source of historical images in the coastal region of Ghana.

The use of continuous time series of satellite images, which provide consistently repeated observations of the earth's surface, can help improve the accuracy of mangroves estimates in Ghana [@Lambin-and-Strahlers-1994; @Verbesselt-et-al-2010; @Zhu-and-Woodcock-2014]. However, differences due to remote sensing artifacts in these multitemporal images may cause error in comparative image analysis. Change detection analysis, therefore, should involve comparable data (i.e., similar sensors, similar radiometric and spatial resolutions, similar viewing geometries, similar image anniversaries). 
The reliability of comparative image analysis for change detection such as post-classification, temporal image differencing, temporal image ratioing, depends on various environmental and atmospheric factors and the accuracy of their resulting change detection analysis depends on the accuracy of the input data. Error in the input images is likely to have compounding effects in the change detection results. These approach, however, are intuitive, simple to implement, and provide quantitative change estimates across a variety of LULC. Yet, for these conventional methods to effectively support any operational application, data input should be comparable across the spatial and temporal domains of the analysis [@Lillesand-et-al-2015].

Multitemporal image analysis should account for potential sources of uncertainty such as imperfection of cloud masking algorithms, sensor differences, and other ephemeral changes induced by erratic events that could introduce error in the classifications [@Zhu-and-Woodcock-2014]. Multi-date images should be of the same period to account for differences in sun angle and temporal variability in LULC. Post-classifications for change detection should account for all major LULC while considering reference data (i.e., training and validation data) over areas of persistent LULC. Besides the requirements for training and validation for LULC classifications, it is important to ensure the comparability of the LULC classifications [@Lillesand-et-al-2015; @Zhu-and-Woodcock-2014]. This can be achieved by either using a single classifier that is consistent with all classifications considered or ensuring that throughputs and accuracies are comparable when using multiple classifiers. These basics requirements have strong implications for the accuracy of change detection analysis that rely on multitemporal image classification. These requirements are often difficult to satisfy due to the lack of ideal satellite images.

While Landsat collections are currently the best publicly available source of data that provide a reasonable and consistent synoptic estimate of land surface state within acceptable spatial resolution for studying changes in mangrove vegetation, it is virtually impossible to obtain the ideal pair of Landsat images in the coastal regions of Ghana. In these areas, the persistence of thick cloud layers reduces the number of usable images [@Rubin-et-al-1999]. Consequently, these images are not always available at the nominal revisit period of 16 days for every pixel.

To address the issues mentioned above, we propose an approach that exploit the landscape phenology to reconstruct comparable data that account for the temporal behaviour of LULC. To ensure data comparability we masked out cloudy pixels and harmonized all available images across different Landsat sensors. To reconstruct gap-free continuous times series we used time-series statistical models to remove the noisy data, fill in missing data and recover the main signal. Based on this signal, we used unsupervised classification to identify areas of persistent LULC. To provide temporally consistent reference data for accurate multi-date classifications and change detection, we tailored our reference data to these persistent LULC. In doing so, the approach accounts for major sources of bias and limits the chance of propagating classification errors into change detection results.

Our goal is to provide consistent up-to-date estimates of mangrove vegetation dynamics of Ghana based on continuous time series of remotely-sensed data acquired between 2000 and 2020. We used all available Tiers 1 and Tiers 2 Landsat data available in Google Earth Engine cloud computing platform to classify the existing LULC of Ghana into 17 categories and study the dynamics of mangrove vegetation.

@@

Bordered by Côte d'Ivoire to the west, Burkina Faso to the north and Togo to the east, Ghana is located on the West African shore of Atlantic Ocean. The country has a rich body of fresh, and salty waters among which the major ones include the red Volta, the Black Volta and the White Volta rivers that originate in Burkina Faso to flow into Lake Volta and its estuary (Figure \ref{fig:fig1}). Ghana has a coastline of about 550 km and over 100 estuaries and lagoons [@Levy-et-al-2015].

As of 2010, the population of Ghana is estimated at nearly 25 million, with an annual population growth rate of 3.1 per annum [@Ghana-Statistical-Services-2012]. The coastal regions (e.g., Accra, Cape Coast) have the highest population density [@Addae-and-Oppelt-2019]. These coastal regions experience two rainy seasons between March to July and September to November, with an annual average temperature of about 26.8 $^\circ$C. The monthly temperature ranges from 24.7 $^\circ$C in August to 33 $^\circ$C in March.

While Landsat remained the publicly available source of data that provides the best compromise between spatial and temporal resolutions in Ghana until Sentinel data become available in 2015, the persistence of cloud poses an important problem for optical remote sensing application [@Ashiagbor-et-al-2021]. The chance of obtaining a clear observation at a given pixel rarely exceed 0.5, particularly in the coastal region (Figure \ref{fig:fig1}). This limits the amount of usable data [@Ashiagbor-et-al-2021], making it difficult to map the entire landscape. This is a region of intensive LULC changes due to anthropogenic activities. Most of these changes, however, are gradual, meaning that the phenology of LULC remains more or less stable over a relatively short period (e.g., 2 to 3 years). For pixels where remote sensing data are available, this phenology can be easily described using time series statistical models.

@@

@@

To consistently capture the phenology of LULC and provide accurate estimates of mangroves in Ghana, we considered all Landsat Tiers 1 and Tiers 2 collections, available from Google Earth Engine cloud computing platform [@Gorelick-et-al-2017; @Wulder-et-al-2019], acquired between 1999 and 2020. We used a total number of 6409 Landsat tiles, each of which composed of 6 bands covering the 3 visible bands, the near infrared band, and the 2 short wave infrared bands. We considered the blue (0.45 - \SI{0.52}{\micro\metre}), the green (0.52 - \SI{0.60}{\micro\metre}), the red (0.63 - \SI{0.69}{\micro\metre}), the near infrared or NIR (0.77 - \SI{0.90}{\micro\metre}), the shortest wave infrared or SWIR1 (1.55 - \SI{1.75}{\micro\metre}) and the longest wave infrared or SWIR2 (2.09 - \SI{2.35}{\micro\metre}) Landsat channels to account for different frequency ranges along the electromagnetic spectrum in relation to differences in the reflectance of LULC. These atmospherically corrected data are suitable for LULC analysis since the surface reflectance values are comparable to those measured on the ground. They are also good candidates for change detection because they account for major atmospheric factors such as clouds, cloud shadow and aerosol scattering.

@@

@@

We conducted extensive field survey to identify 17 LULC classes based on which we systematic ally collected reference data for training and validation of LULC classifications. To provide a comprehensive LULC nomenclature, we defined most of these LULC classes (Table \ref{tab1}) based on the standard IGBP and FAO LULC classification systems [@Di-Gregorio-et-al-2016; @FRA-2000]. 
The first step in the collection of the reference data consisted of identifying regions of dominant LULC and collecting large polygons over these regions. These reference polygons are distributed over the area of interest in a way that captures the essential LULC variability (both between and within class variability). The second step consisted of identifying pixels of persistent LULC within each reference polygon and extracting the reflectance values of these pixels from the multiband image. Pixels of persistent LULC are pixels whose class remains unchanged across the reference periods considered. We used these persistent pixels to ensure that the reflectance values used for training and validation are consistent with the phenology of their corresponding LULC across the reference periods.

@@

```{r tab1, echo=FALSE}
cbind(
  `Standard class` = c( 
    'Water bodies',
    'Closed forests',
    'Open forests',
    'Woody savannas',
    'Savannas',
    'Closed shrublands',
    'Open shrublands',
    'Grasslands',
    'Permanent wetlands',
    'Croplands',
    'Urban and built-up lands',
    'Cropland / natural vegetation mosaics',
    'Barren',
    'Mangroves',
    'Salt mines',
    'Tree plantations',
    'Regularly Flooded Vegetation'
  ),

  `Class description` = c(
    'Natural and artificial Water bodies such as ocean, rivers, and other reservoirs containing fresh or salty water for the most period of the year.',
    'Lands dominated by trees with a percent cover greater than 70 % during the entire period of the year.',
    'Lands dominated by trees with a percent cover between 60 and 70 % during the entire period of the year.',
    'Lands with herbaceous and other understory systems, and with forest canopy cover between 30% and 60%. The forest cover height exceeds 2 m.',
    'Lands with herbaceous and other understory systems, and with forest canopy cover between 10% and 30%. The forest cover height exceeds 2 m.',
    'Lands with woody vegetation less than 2 m tall and with shrub canopy cover > 60%. The shrub foliage can be either evergreen or deciduous.',
    'Lands with woody vegetation less than 2 m tall and with shrub canopy cover between 10% and 60%. The shrub foliage can be either evergreen or deciduous.',
    'Lands with herbaceous types of cover. Tree and shrub cover is less than 10%.\n',
    'Lands with a permanent mixture of fresh water and herbaceous or woody vegetation.',
    'Lands covered with temporary crops followed by harvest and a bare soil period (e.g., single and multiple cropping systems).',
    'Land covered by buildings and other man-made structures. This class includes all concrete cover, such as roads.',
    'Lands with a mosaic of croplands, forests, shrubland, and grasslands in which no one component comprises more than 60% of the landscape.',
    'Lands with exposed soil, sand, rocks, or snow and never have more than 10% vegetated cover during any time of the year.',
    'Lands with a permanent mixture of brackish water and herbaceous or woody vegetation.',
    'Land with water of high salt concentration where the salt naturally emerges or is artificially extracted from evaporite formations.',
    'Land with perennial crop in the form of artificial forests. These are mostly represented by palm trees in the coastal region of Ghana.',
    'Land transitioning between terrestrial and fresh water zones with sufficient moisture for the development of near evergreen vegetation.'
  ),

  `Map legend` = c(
    'Waters',
    'Closed forests',
    'Open forests',
    'Woody savannas',
    'Savannas',
    'Closed shrublands',
    'Open shrublands',
    'Grasslands',
    'Wetlands',
    'Croplands',
    'Built-up',
    'Mosaics',
    'Barren',
    'Mangroves',
    'Salt mines',
    'Tree plantations',
    'Riparian vegetation'
  )
) %>%
  kable(format = "latex", longtable = FALSE, booktabs = TRUE, digits = 2, 
        escape = TRUE, format.args = list(big.mark = ","),
        linesep = '\\addlinespace',
        align = c('l', 'l', 'r'),
        caption = '\\label{tab1}Categories of land use and land cover used for an image classification in Ghana.') %>% 
  row_spec(0,bold=TRUE) %>% 
  # row_spec(14,bold=TRUE) %>%
  # row_spec(1:17, hline_after = TRUE) %>%
  # column_spec(1, width = "20em") %>%
  # column_spec(2, width = "35em") %>%
  # column_spec(3, width = "10em") %>%
  kable_styling(latex_options = c("repeat_header", "scale_down", "HOLD_position")) %>%
  footnote(general = "Regularly Flooded Vegetation, closed forests and open forests are based on FAO LULC classification system. Tree plantations are mostly represented by palm trees. The remaining classes are based on IGBP land cover classification system except for Mangroves which is the class of interest and Salt mines which is relevant in regards to mangroves vegetation dynamics in Ghana. Perennial woody crops are classified as either tree plantations (mostly palm trees) or the appropriate forest or shrub land cover type.", threeparttable = TRUE)
```

@@

Prior to image classifications, we conducted a number of tasks to minimize the effects of external factors on the resulting LULC classification. These image pre-processing tasks involved several steps including the masking of cloudy pixels, the scaling of reflectance values across the different Landsat sensors, the removal of noisy data, gap filling, and the processing of the phenological signal.

@@

We used the quality assessment band provided along with the surface reflectance product to mask all cloudy pixels. Despite the similarity of Landsat sensors, data from different sensors present a slight difference which can be relevant for analysis involving multiple sensors such as cross-sensor time series, temporal image composite, or gap filling. Therefore, we scaled the reflectance to the value range of OLI sensor using the harmonisation coefficients provided by @Roy-et-al-2016.

@@

Relatively high RMSE are rare. The model appears to have achieved the best performance over the short-wave infrared channels with the shortest wave infrared being the best predicted. The model performance over the shortest wave infrared appears to be comparable to that over the near infrared. The model appears to have achieved, more or less similarly, the lowest performance over the visible bands where difference across tiles becomes more perceptible. In general, the model performance is relatively poor over the coastal regions and areas of dense vegetation. While water bodies appear to be essentially associated with the lowest RMSE, this RMSE appears to slightly increase over riparian vegetation. In general, the performance of the model appears to increase with wavelength and seasonality.

>>>>

---
title: Estimating the dynamics of Ghanaian mangrove cover using phenological metrics derived from synoptic Landsat observations
author:
  - name: Issoufou Liman Harou
    email: issoufoul@gmail.com
    affiliation: ICRAFGOV,KUENV
    correspondingauthor: true
  - name: Julliet Inyele 
    email: inyelejuli@gmail.com
    affiliation: KUGEO
  - name: Peter Minang  
    email: A.Minang@cgiar.org
    affiliation: ICRAFGOV
  - name: Lalisa Duguma
    email: l.duguma@cgiar.org
    affiliation: ICRAFGOV
address:
  - code: ICRAFGOV
    address: "Governance and Landscapes, World Agroforestry (ICRAF), UN Avenue, Gigiri, Nairobi, Kenya, P.O. Box: 30677-00100"
  - code: KUENV
    address: "Department of Environmental Sciences, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O. Box: 43844-00100"
  - code: KUGEO
    address: "Department of Geography, Kenyatta University, Kenya Drive, Nairobi, Kenya, P.O. Box: 43844-00100"
abstract: |
  Remote sensing analysis and narratives from Ghana (West Africa) reported evidence of mangrove degradation across the southern regions of the country. Despite the importance of mangrove forests, the coverage of earth observation satellites and the recent advances in computing power for processing spatial data, spatially-explicit estimates of mangrove vegetation cover are lacking. In these regions, optical sensors onboard Landsat satellites provide a great deal of historical remote sensing data for studying land surface dynamics.  While these sensors provide reasonable spatial, temporal and spectral resolutions for studying the dynamics of mangrove vegetation cover, a number of remote sensing artifacts make it difficult to derive a complete land cover map.  In particular, the persistence of cloud cover limits the number of usable images. To understand the dynamics of mangrove forests, we identified 17 land cover classes, harmonized the Landsat scenes across sensors, used time series analysis to improve cloud detection and recover the relevant signal encoding the temporal signature of land cover. Based on the seasonal behavior of the land cover classes, the persistence of cloud cover limiting the number of usable images and the temporal resolution of Landsat satellite, we mosaiced the images within 30 days of their acquisition dates. Using the fitted image time series, we computed unsupervised classifications which we used along with ground control polygons to generate training and validation data for land cover classifications and change detection. The analysis of the root mean square error indicates that the accuracy of the fitted image time series increases with band wavelength and land cover seasonality. The accuracy assessment indicates an overall accuracy between 0.94 and 0.95 and a Kappa between 0.93 and 0.95 for the different land cover classifications. Mangrove forests have declined in some areas, whereas they have increased in others. Mangrove area, which has decreased between 2000 and 2010, has increased between 2010 and 2020. The gains experienced between 2010 and 2020, however, did not compensate for the lost the land cover experienced between 2000 and 2010. We identified four top most important land cover involved in mangroves conversion including forests with a high rate, croplands and natural vegetation mosaics and tree plantations with moderate rate and wetlands with a low rate. Better mangrove restoration and conservation are required to fully reverse the trend of mangrove decline.
keywords: 
  - Cloud Mask
  - Google Earth Engine
  - Landsat Data
  - Land Cover Classification
  - Mangrove
journal: "Remote Sensing of Environment"
date: "`r Sys.Date()`"
# classoption: preprint, 3p, authoryear
classoption: 12pt,oneside,preprint,3p,authoryear,times
bibliography: bibliography.bib
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/remote-sensing-of-environment
csl: remote-sensing-of-environment.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: natbib
link-citations: yes
linkcolor: blue
header-includes:
  - \input{includes.tex}
note: Replication files are available on the author's Github account (http://github.com/Issoufou-Liman).
---

@@

Despite the importance of mangroves, some studies reported that mangrove vegetation cover has been alarmingly declining in many areas of Ghana for the last few decades [@Asante-et-al-2017]. @Armah-et-al-2009 estimated this decline to range between 20 and 30% across the coast of Atlantic Ocean over the past 3 decades. @FAO-2007 reported a mangrove vegetation loss of 0.5 million ha (13.8 % of the total mangrove areas) in 25 years. According to @Rubin-et-al-1999, almost 2/3 of the Volta estuary mangroves have been lost since 1973, leaving only a few patches within a buffer of 15 km. @Darkwa-and-Smardon-2010 noticed an increased cutting of mangroves vegetation in Fosu Lagoon. @FAO-2007 estimated the decline of mangrove areas in Ghana at 24% between 1980 to 2005. In other areas, however, other studies reported an appreciable increase in mangrove vegetation cover following restoration efforts by the locals. @Awo-et-al-2014 reported a noticeable increase in mangrove cover between 1986 and 2002 in Anyanui (Keta Lagoon). @Feka-2015 noted that despite the reduction in mangrove area in Ghana, an estimated area of 68 ha of mangroves have been restored from 1980 to 2006.

It becomes clear from the available information that a country-wide assessment of mangrove vegetation in Ghana is needed. A holistic assessment of all major Land Use Land Cover (LULC) would help to understand the dynamics of mangrove vegetation and guide mangrove conservation policy in the study area. This would help identify areas of increase and decrease in mangrove vegetation, understand the trend in LULC change, identify areas that require immediate action and those areas where policy interventions can achieve the greatest impact. To our knowledge, country-wide assessments of mangrove vegetation that deliver this crucial information for mangrove conservation are lacking. The most recent and comprehensive ones (e.g., @CILSS-2016; @Spalding-et-al-1997; @Spalding-et-al-2010) have either coarse spatial resolution or have important information gap to be suitable for this purpose. This information gap is mainly due to the prevalence of cloud cover [@Ashiagbor-et-al-2021] that makes it virtually impossible to derive a complete land cover map from a single acquisition [@Rubin-et-al-1999]. Clouds deserved a standalone class in @CILSS-2016 because none of the best available image was cloud-free enough to allow a comprehensive post-classification interpolation. Based on our analysis of clouds identified by Fmask algorithm, at least half of the available Landsat images acquired between 2000 and 2020 presents missing observations over the coastal regions of Ghana. There is a need for approaches that could comprehensively account for these gaps when working with optical remote sensing data which are currently the best available source of historical images in the coastal region of Ghana.

The use of continuous time series of satellite images, which provide consistently repeated observations of the earth's surface, can help improve the accuracy of mangroves estimates in Ghana [@Lambin-and-Strahlers-1994; @Verbesselt-et-al-2010; @Zhu-and-Woodcock-2014]. However, differences due to remote sensing artifacts in these multitemporal images may cause error in comparative image analysis. Change detection analysis, therefore, should involve comparable data (i.e., similar sensors, similar radiometric and spatial resolutions, similar viewing geometries, similar image anniversaries). 
The reliability of comparative image analysis for change detection such as post-classification, temporal image differencing, temporal image ratioing, depends on various environmental and atmospheric factors and the accuracy of their resulting change detection analysis depends on the accuracy of the input data. Error in the input images is likely to have compounding effects in the change detection results. These approach, however, are intuitive, simple to implement, and provide quantitative change estimates across a variety of LULC. Yet, for these conventional methods to effectively support any operational application, data input should be comparable across the spatial and temporal domains of the analysis [@Lillesand-et-al-2015]. Multitemporal image analysis should account for potential sources of uncertainty such as imperfection of cloud masking algorithms, sensor differences, and other ephemeral changes induced by erratic events that could introduce error in the classifications [@Zhu-and-Woodcock-2014]. Multi-date images should be of the same period to account for differences in sun angle and temporal variability in LULC. Post-classifications for change detection should account for all major LULC while considering reference data (i.e., training and validation data) over areas of persistent LULC. Besides the requirements for training and validation for LULC classifications, it is important to ensure the comparability of the LULC classifications [@Lillesand-et-al-2015; @Zhu-and-Woodcock-2014]. This can be achieved by either using a single classifier that is consistent with all classifications considered or ensuring that throughputs and accuracies are comparable when using multiple classifiers. These basics requirements have strong implications for the accuracy of change detection analysis that rely on multitemporal image classification. These requirements are often difficult to satisfy due to the lack of ideal satellite images. While Landsat collections are currently the best publicly available source of data that provide a reasonable and consistent synoptic estimate of land surface state within acceptable spatial resolution for studying changes in mangrove vegetation, it is virtually impossible to obtain the ideal pair of Landsat images in the coastal regions of Ghana. In these areas, the persistence of thick cloud layers reduces the number of usable images [@Rubin-et-al-1999]. Consequently, these images are not always available at the nominal revisit period of 16 days for every pixel.

To address the issues mentioned above, we propose an approach that exploit the landscape phenology to reconstruct comparable data that account for the temporal behaviour of LULC. To ensure data comparability we masked out cloudy pixels and harmonized all available images across different Landsat sensors. To reconstruct gap-free continuous times series we used time-series statistical models to remove the noisy data, fill in missing data and recover the main signal. Based on this signal, we used unsupervised classification to identify areas of persistent LULC. To provide temporally consistent reference data for accurate multi-date classifications and change detection, we tailored our reference data to these persistent LULC. In doing so, the approach accounts for major sources of bias and limits the chance of propagating classification errors into change detection results. Our goal is to provide consistent up-to-date estimates of mangrove vegetation dynamics of Ghana based on continuous time series of remotely-sensed data acquired between 2000 and 2020. We used all available Tiers 1 and Tiers 2 Landsat data available in Google Earth Engine cloud computing platform to classify the existing LULC of Ghana into 17 categories and study the dynamics of mangrove vegetation.

@@

Bordered by Côte d'Ivoire to the west, Burkina Faso to the north and Togo to the east, Ghana is located on the West African shore of Atlantic Ocean. The country has a

@@

\noindent rich body of fresh, and salty waters among which the major ones include the red Volta, the Black Volta and the White Volta rivers that originate in Burkina Faso to flow into Lake Volta and its estuary (Figure \ref{fig:fig1}). Ghana has a coastline of about 550 km and over 100 estuaries and lagoons [@Levy-et-al-2015].

As of 2010, the population of Ghana is estimated at nearly 25 million, with an annual population growth rate of 3.1 per annum [@Ghana-Statistical-Services-2012]. The coastal regions (e.g., Accra, Cape Coast) have the highest population density [@Addae-and-Oppelt-2019]. These coastal regions experience two rainy seasons between March to July and September to November, with an annual average temperature of about 26.8 $^\circ$C. The monthly temperature ranges from 24.7 $^\circ$C in August to 33 $^\circ$C in March. While Landsat remained the publicly available source of data that provides the best compromise between spatial and temporal resolutions in Ghana until Sentinel data become available in 2015, the persistence of cloud poses an important problem for optical remote sensing application [@Ashiagbor-et-al-2021]. The chance of obtaining a clear observation at a given pixel rarely exceed 0.5, particularly in the coastal region (Figure \ref{fig:fig1}). This limits the amount of usable data [@Ashiagbor-et-al-2021], making it difficult to map the entire landscape. This is a region of intensive LULC changes due to anthropogenic activities. Most of these changes, however, are gradual, meaning that the phenology of LULC remains more or less stable over a relatively short period (e.g., 2 to 3 years). For pixels where remote sensing data are available, this phenology can be easily described using time series statistical models.

@@

To consistently capture the phenology of LULC and provide accurate estimates of mangroves in Ghana, we considered all Landsat Tiers 1 and Tiers 2 collections, available

@@

\noindent from Google Earth Engine cloud computing platform [@Gorelick-et-al-2017; @Wulder-et-al-2019], acquired between 1999 and 2020. We used a total number of 6409 Landsat tiles, each of which composed of 6 bands covering the 3 visible bands, the near infrared band, and the 2 short wave infrared bands. We considered the blue (0.45 - \SI{0.52}{\micro\metre}), the green (0.52 - \SI{0.60}{\micro\metre}), the red (0.63 - \SI{0.69}{\micro\metre}), the near infrared or NIR (0.77 - \SI{0.90}{\micro\metre}), the shortest wave infrared or SWIR1 (1.55 - \SI{1.75}{\micro\metre}) and the longest wave infrared or SWIR2 (2.09 - \SI{2.35}{\micro\metre}) Landsat channels to account for different frequency ranges along the electromagnetic spectrum in relation to differences in the reflectance of LULC. These atmospherically corrected data are suitable for LULC analysis since the surface reflectance values are comparable to those measured on the ground. They are also good candidates for change detection because they account for major atmospheric factors such as clouds, cloud shadow and aerosol scattering.

@@

We conducted extensive field survey to identify 17 LULC classes based on which we systematic ally collected reference data for training and validation of LULC classifications. To provide a comprehensive LULC nomenclature, we defined most of these LULC classes (Table \ref{tab1}) based on the standard IGBP and FAO LULC classification systems [@Di-Gregorio-et-al-2016; @FRA-2000].

```{r tab1, echo=FALSE}
cbind(
  `Standard class` = c( 
    'Water bodies',
    'Closed forests',
    'Open forests',
    'Woody savannas',
    'Savannas',
    'Closed shrublands',
    'Open shrublands',
    'Grasslands',
    'Permanent wetlands',
    'Croplands',
    'Urban and built-up lands',
    'Cropland and natural vegetation mosaics',
    'Barren',
    'Mangroves',
    'Salt mines',
    'Tree plantations',
    'Regularly Flooded vegetation'
  ),

  `Class description` = c(
    'Natural and artificial Water bodies such as ocean, rivers, and other reservoirs containing fresh or salty water for the most period of the year.',
    'Lands dominated by trees with a percent cover greater than 70 % during the entire period of the year.',
    'Lands dominated by trees with a percent cover between 60 and 70 % during the entire period of the year.',
    'Lands with herbaceous and other understory systems, and with forest canopy cover between 30% and 60%.The forest cover height exceeds 2 m.',
    'Lands with herbaceous and other understory systems, and with forest canopy cover between 10% and 30%. The forest cover height exceeds 2 m.',
    'Lands with woody vegetation less than 2 m tall and with shrub canopy cover > 60%. The shrub foliage can be either evergreen or deciduous.',
    'Lands with woody vegetation less than 2 m tall and with shrub canopy cover between 10% and 60%. The shrub foliage can be either evergreen or deciduous.',
    'Lands with herbaceous types of cover. Tree and shrub cover is less than 10%.',
    'Lands with a permanent mixture of fresh water and herbaceous or woody vegetation.',
    'Lands covered with temporary crops followed by harvest and a bare soil period (e.g., single and multiple cropping systems).',
    'Land covered by buildings and other man-made structures. This class includes all concrete cover, such as roads.',
    'Lands with a mosaic of croplands, forests, shrubland, and grasslands in which no one component comprises more than 60% of the landscape.',
    'Lands with exposed soil, sand, rocks, or snow and never have more than 10% vegetated cover during any time of the year.',
    'Lands with a permanent mixture of brackish water and herbaceous or woody vegetation.',
    'Land with water of high salt concentration where the salt naturally emerges or is artificially extracted from evaporite formations.',
    'Land with perennial crop in the form of artificial forests. These are mostly represented by palm trees in the coastal region of Ghana.',
    'Land transitioning between terrestrial and fresh water zones with sufficient moisture for the development of near evergreen vegetation.'
  ),

  `Map legend` = c(
    'Waters',
    'Closed forests',
    'Open forests',
    'Woody savannas',
    'Savannas',
    'Closed shrublands',
    'Open shrublands',
    'Grasslands',
    'Wetlands',
    'Croplands',
    'Built-up',
    'Mosaics',
    'Barren',
    'Mangroves',
    'Salt mines',
    'Tree plantations',
    'Riparian vegetation'
  )
) %>%
  kable(format = "latex", longtable = FALSE, booktabs = TRUE, 
        escape = TRUE, format.args = list(big.mark = ","),
        linesep = '\\addlinespace',
        align = c('l', 'l', 'r'),
        caption = '\\label{tab1}Categories of land use and land cover used for an image classification in Ghana.') %>% 
  row_spec(0,bold=TRUE) %>% 
  # row_spec(14,bold=TRUE) %>%
  # row_spec(1:17, hline_after = TRUE) %>%
  # column_spec(1, width = "20em") %>%
  column_spec(2, width = "35em") %>%
  # column_spec(3, width = "10em") %>%
  kable_styling(latex_options = c("repeat_header", "scale_down", "HOLD_position")) %>%
  footnote(general = "Regularly Flooded Vegetation, closed forests and open forests are based on FAO LULC classification system. Tree plantations are mostly represented by palm trees. The remaining classes are based on IGBP land cover classification system except for Mangroves which is the class of interest and Salt mines which is relevant in regards to mangroves vegetation dynamics in Ghana. Perennial woody crops are classified as either tree plantations (mostly palm trees) or the appropriate forest or shrub land cover type.", threeparttable = TRUE)
```

The first step in the collection of the reference data consisted of identifying regions of dominant LULC and collecting large polygons over these regions. These reference polygons are distributed over the area of interest in a way that captures the essential LULC variability (both between and within class variability). The second step consisted of identifying pixels of persistent LULC within each reference polygon and extracting the reflectance values of these pixels from the multiband image. Pixels of persistent LULC are pixels whose class remains unchanged across the reference periods considered. We used these persistent pixels to ensure that the reflectance values used for training and validation are consistent with the phenology of their corresponding LULC across the reference periods.

@@

@@

Prior to image classifications, we conducted a number of tasks to minimize the effects of external factors on the resulting LULC classification. These image pre-processing tasks involved several steps including the masking of cloudy pixels, the scaling of reflectance values across the different Landsat sensors, the removal of noisy data, gap filling, and the processing of the phenological signal. We used the quality assessment band provided along with the surface reflectance product to mask all cloudy pixels. Despite the similarity of Landsat sensors, data from different sensors present a slight difference which can be relevant for analysis involving multiple sensors such as cross-sensor time series, temporal image composite, or gap filling. Therefore, we scaled the reflectance to the value range of OLI sensor using the harmonisation coefficients provided by @Roy-et-al-2016.

@@

@@
